---
title: "Assessment"
output: html_document
date: "2025-10-31"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Exploring the relationship between number of cholesterol-lowering drug prescriptions and socioeconomic group in Health Boards across Scotland 2024-2025.

Downlaod libraries
```{r}
library(tidyverse)
library(janitor) # cleaning data
library(gt) # tables
library(here) # directory structure (will be useful later)
```

Load data:
I chose to incorporate data from the end of 2024 and beginning of 2025 to give a representative account of the most recent data available of Scotland.
```{r}
#health board data for july-december 2024 and jan-june 2025
hb_data_2025 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9de908b3-9c28-4cc3-aa32-72350a0579d1/download/hb_pitc2025_01_06.csv") %>% 
  clean_names() %>% 
  select(hbt, bnf_item_description, paid_quantity)

hb_data_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/f3b9f2e2-66c0-4310-9b8e-734781d2ed0a/download/hb_pitc2024_07_12-1.csv") %>% 
  clean_names() %>% 
  select(hbt, bnf_item_description, paid_quantity)

#health board names associated with each HB code
health_board_names <- read_csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% 
  clean_names() %>% 
  select(hb, hb_name)

#Scotland population data
library(readr)
population_data <- read_csv(here( "data", "UV102a_age_health_board_census.csv"), skip = 10) 

#Scotland National Statistics Socio-economic Classification (NS-SeC) of Household Reference Person
se_data <- read_csv(here( "data", "table_2025-11-12_18-21-02.csv"), skip = 10) 
  
```

Join 2024 and 2025 data:
```{r}
hb_all <- bind_rows(hb_data_2024, hb_data_2025)
```

Edit population data:
```{r}
library(readr)
population_data <- read_csv(here("data/UV102a_age_health_board_census.csv"), skip = 10) %>% 
  rename(Spare = "...6",
         hb_name = "Health Board Area 2019",
         hb_population = Count) %>% 
  filter(Age == "All people" & Sex == "All people") %>% 
  select(hb_name, hb_population) %>% 
  mutate(hb_name = paste("NHS", hb_name))
```

Grouping SE data into low, middle and high SE class:
```{r}
se_data <- read_csv(here( "data", "table_2025-11-12_18-21-02.csv"), skip = 10) %>%
  #grouping all 15 levels of socioeconomic group to make data more easily digestable without exclusing any groups of data groups to maintain representation.
  mutate(upper = rowSums(across(c("L1: Employers in large establishments", "L2: Higher managerial and administrative occupations", "L3: Higher professional occupations")), na.rm = TRUE),
         middle = rowSums(across(c("L4: Lower professional and higher technical occupations", "L5: Lower managerial and administrative occupations", "L6: Higher supervisory occupations", "L7: Intermediate occupations", "L8: Employers in small establishments", "L9: Own account workers")), na.rm = TRUE), 
         low = rowSums(across(c("L10: Lower supervisory occupations", "L11: Lower technical occupations", "L12: Semi-routine occupations", "L13: Routine occupations", "L14.1: Never worked", "L14.2: Long-term unemployed", "L15: Full-time students"))), na.rm = TRUE) %>% 
  select("NS-SeC of Household Reference Person", "All occupied households", "upper", "middle", "low") %>% 
  clean_names() %>% 
#removing NA values 
  drop_na() %>% 
  mutate(
    upper = as.numeric(upper),
    middle = as.numeric(middle),
    low = as.numeric(low),
    all_occupied_households = as.numeric(all_occupied_households)) %>% 
  # finding percentages for each socioeconomic group
  mutate(upper_pct = upper/all_occupied_households*100,
         middle_pct = middle/all_occupied_households*100,
         low_pct = low/all_occupied_households*100) %>% 
  pivot_longer(cols = c(upper_pct, middle_pct, low_pct), 
               names_to = "socioeconomic_group",
               values_to = "households")
#plotting SE data
se_data_plot <- se_data %>% 
  ggplot(aes(x = ns_se_c_of_household_reference_person, y = households, fill = socioeconomic_group)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "National Statistics Socio-economic Classification of Households across Health Boards in Scotland",
       x = "Health Board",
       y = "Number of Households")
 se_data_plot 
```
This plot depicts the socioeconomic groups categorised into upper, middle, and low across Health Boards in Scotland. There does not seem to be a huge discrepancy - Lothian has the highest percentage of households in the upper percentile, with Western Isles and Dumfries and Galloway having the lowest in the upper percentile of socioeconomic group. We aim to investigate whether or not there is a correlation between socioeconomic group and number of cholesterol-lowering drug prescriptions among Scottish Health Boards.








