---
title: "Exploring the relationship between number of cholesterol-lowering drug prescriptions and socioeconomic group in Health Boards across Scotland 2024-2025."
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
date: "2025-11-27"

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 5) # making figures full width for better visualisation
```

# Introduction

A previous study in 2014, found that diabetes and cardiovascular disease risk factors are more prevalent in deprived than affluent populations in Scotland. Such factors include smoking status, BMI, cholesterol levels, and age (Jones et al., 2014). Recent reports urge the need for increased recognition of socioeconomic deprivation when dealing with risk estimates for cardiovascular disease, to ultimately reduce inequalities in healthcare provision (Kimenai et al., 2022). This is an increasing issue, globally, but Scotland has been particularly recognised in this instance. In this report, we aim to investigate if there is in fact a link, by looking into the correlation between number of cholesterol-lowering drug prescriptions and socioeconomic (SE) group across health boards in Scotland in 2024 to 2025. 

## Loading libraries

```{r}
library(tidyverse)
library(janitor) # cleaning data
library(gt) # tables
library(here)
library(gtExtras)
library(kableExtra)
library(plotly) # interactive plots
```

## Loading data

I joined data from the end of 2024 and beginning of 2025 to give a representative account of the most recent data available of Scotland. In combination with this health board data, I also loaded population so I could calculate number of prescriptions per head, as well as data from the Scotland census including Scotland national statistics socioeconomic classification (NS-SeC) of household reference person.

All data used in this report can be accessed via links provided at the bottom of the table in section 5. 

```{r}
# health board data for January - June 2025
hb_data_2025 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9de908b3-9c28-4cc3-aa32-72350a0579d1/download/hb_pitc2025_01_06.csv") %>% 
  clean_names() %>% 
  select(hbt, bnf_item_description, paid_quantity)

# health board data for July - December 2024
hb_data_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/f3b9f2e2-66c0-4310-9b8e-734781d2ed0a/download/hb_pitc2024_07_12-1.csv") %>% 
  clean_names() %>% 
  select(hbt, bnf_item_description, paid_quantity)

# joining health board data from 2024 and 2025.
hb_all <- bind_rows(hb_data_2024, hb_data_2025)

# health board names associated with each HB code
health_board_names <- read_csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% 
  clean_names() %>% 
  select(hb, hb_name)

# Scotland census population data
population_data <- read_csv(here( "data", "UV102a_age_health_board_census.csv"), skip = 10) %>% 
  rename(Spare = "...6", #selecting only HB name and HB population
         hb_name = "Health Board Area 2019",
         hb_population = Count) %>% 
  filter(Age == "All people" & Sex == "All people") %>% 
  select(hb_name, hb_population) %>% 
  mutate(hb_name = paste("NHS", hb_name))

# Scotland National Statistics Socio-economic Classification (NS-SeC) of Household Reference Person
se_data <- read_csv(here( "data", "table_2025-11-12_18-21-02.csv"), skip = 10) 
  
```

# Socioeconomic data: Plot 1 

Plot 1 aims to investigate and organise the populations of Scottish health boards into upper, middle, and low SE group. This will allow us to find the percentage of each, per health board. There are 15 levels of data, which have been categorised for improved visualisation. Further, the data is filtered for only the upper percentile of each health board population so that we later directly compare only the percentage of the population of a given health board in the upper socioeconomic percentile, with number of cholesterol-lowering prescriptions. 

## Data wrangling

Grouping SE data into upper, middle, and low SE class:
```{r}
se_data <- read_csv(here( "data", "table_2025-11-12_18-21-02.csv"), skip = 10) %>%
  # grouping all 15 levels of socioeconomic group to make data more easily digestible without excluding any data groups to maintain representation of Scotland's population as a whole.
  mutate(upper = rowSums(across(c("L1: Employers in large establishments", "L2: Higher managerial and administrative occupations", "L3: Higher professional occupations")), na.rm = TRUE),
         middle = rowSums(across(c("L4: Lower professional and higher technical occupations", "L5: Lower managerial and administrative occupations", "L6: Higher supervisory occupations", "L7: Intermediate occupations", "L8: Employers in small establishments", "L9: Own account workers")), na.rm = TRUE), 
         low = rowSums(across(c("L10: Lower supervisory occupations", "L11: Lower technical occupations", "L12: Semi-routine occupations", "L13: Routine occupations", "L14.1: Never worked", "L14.2: Long-term unemployed", "L15: Full-time students"))), na.rm = TRUE) %>% 
  select("NS-SeC of Household Reference Person", "All occupied households", "upper", "middle", "low") %>% 
  clean_names() %>% 
# removing NA values and converting values to be numeric. This allows us to then find percentage of population so we can directly compare, independent of large population differences between health boards. 
  drop_na() %>% 
    mutate(upper = as.numeric(upper),
    middle = as.numeric(middle),
    low = as.numeric(low),
    all_occupied_households = as.numeric(all_occupied_households)) %>% 
   # finding percentages for each SE group
  mutate(upper_pct = upper/all_occupied_households*100,
         middle_pct = middle/all_occupied_households*100,
         low_pct = low/all_occupied_households*100) %>% 
  pivot_longer(cols = c(upper_pct, middle_pct, low_pct), 
               names_to = "socioeconomic_group",
               values_to = "households") %>% 
  rename(hb_name = ns_se_c_of_household_reference_person) %>% 
  mutate(hb_name = paste0("NHS ", hb_name))  

#filtering for only percentage of households in upper percentile per health board as this is what we are interested in.
se_data_upper <- se_data %>%
  filter(socioeconomic_group == "upper_pct") %>% 
  select(-all_occupied_households, -upper, -middle, -low, -socioeconomic_group) %>% 
  rename(upper_household_pct = households)
```

## Creating a function

```{r}
#writing a function for the theme
  theme_custom1 <- function(base_text_size = 10) {
    theme_minimal(base_size = base_text_size, base_family = "Times New Roman") +
      theme(plot.title = element_text(colour = "#a90c4b", face = "bold"),
                  axis.text.x = element_text(colour = "#a90c4b", angle = 45, hjust = 1),
            axis.text.y = element_text(colour = "#a90c4b"),
            axis.title = element_text(colour = "#a90c4b"),
            plot.background = element_rect(fill = "#ffffe0"),)
    }
```

## Plotting socioeconomic data

In Plot 1, we aim to visualise how the percentage of households in the upper SE group is distributed across Scottish health boards. 

```{r}
# plotting SE data
  se_data_plot <- se_data_upper %>% 
    mutate(hover_text = paste0("Percentage of Households: ", sprintf("%.2f%%", upper_household_pct))) %>% 
  ggplot(aes(x = hb_name, y = upper_household_pct, fill = upper_household_pct, text = hover_text)) +
  geom_col() +
  scale_fill_gradient(low = "sandybrown", high = "#94023f") + # visual indicator of quantity.
  theme_custom1() +
  labs(title = "National Statistics Socio-economic Classification of Households \n across Health Boards in Scotland",
       x = "Health Board",
       y = "Percentage of Households in Upper Socioeconomic Group",
       fill = "% of population in upper percentile")

# interactive plot - allowing hover for exact value.
 se_data_plotly <- plotly::ggplotly(se_data_plot, tooltip = "text") %>% 
   layout(plot_bgcolor = "#ffffe0")
 se_data_plotly 
```

This plot depicts the distribution of the upper socioeconomic group across Health Boards in Scotland. From the plot, it is clear that NHS Lothian has the highest percentage of households in the upper percentile (20.15%), followed by NHS Grampian (15%). In contrast, NHS Western Isles (9.98%) and NHS Dumfries and Galloway (9.71%) have the lowest percentage of households in the upper percentile of socioeconomic group. We now aim to investigate the number of prescriptions of all cholesterol-lowering drugs per head in each health board. This will allow us to make direct comparisons with socioeconomic group distribution, to investigate if there is a relationship.

# Total cholesterol-loweing prescriptions: Plot 2

The number of prescriptions of cholesterol-lowering drugs per head in Health Boards in Scotland in 2024-2025.

## Data wrangling

```{r}
# joining prescription data with health board names.
cl_prescriptions <- hb_all %>% 
  full_join(health_board_names, join_by(hbt == hb)) %>% 
# filter for only cholesterol- lowering drugs indicated by "STATIN" in the name.
   filter(str_detect(bnf_item_description, "STATIN")) %>% 
  group_by(bnf_item_description) %>% 
  select(hbt, bnf_item_description, paid_quantity, hb_name)
```

```{r}
# a visual depiction of all statin drugs in the all health boards in 2024 and 2025, allowing us to more easily identify all of them, without the distraction of repetitions in the original dataset.
cl_prescriptions_visual <- cl_prescriptions %>% 
  distinct(bnf_item_description) %>% 
  arrange(bnf_item_description)
# list of drugs used:
#atorvastatin, fluvastatin, nystatin, pravastatin, rosuvastatin, sandostatin, simvastatin, ecostatin.
```

Drugs were grouped due to the large variation in doses for each drug which makes the number of prescriptions themselves unclear. We are investigating the prescription frequency itself, not the variation of doses. We did grouping so that we could visualise and appropriately investigate the proposed hypothesis.

```{r}
# joining with population data to find prescriptions per head. 
grouped_cl_prescriptions <- cl_prescriptions %>% #'cl' notation stands for cholesterol-lowering
  left_join(population_data, by = c("hb_name" = "hb_name")) %>% 
# extracting and grouping by statin drug group name, independent of dose used.
  mutate(drug_name = str_extract(bnf_item_description, "ATORVASTATIN|FLUVASTATIN|NYSTATIN|PRAVASTATIN|ROSUVASTATIN|SANDOSTATIN|SIMVASTATIN|ECOSTATIN")) %>% 
# finding quantity per head.
  mutate(quantity_per_head = paid_quantity/hb_population) %>% 
  filter(!is.na(drug_name) & !is.na(hb_name)) %>% 
  # filtered for only drug names, independent of dosage.
  group_by(drug_name, hb_name) %>% 
  summarise(total_quantity_per_head = sum(quantity_per_head)) 
  
```

## Creating a vector

```{r}
# assign colors to each statin drug group
drug_colors <- c(
  "ATORVASTATIN" = "indianred",
  "ECOSTATIN" = "snow3", 
  "FLUVASTATIN" = "sandybrown",
  "NYSTATIN" = "seashell",
  "PRAVASTATIN" = "pink3",
  "ROSUVASTATIN" = "salmon",
  "SANDOSTATIN" = "#a90c4b",
  "SIMVASTATIN" = "lightpink"
)
```

## Plotting cholesterol-lowering prescriptions per head 

This plot shows a visualisation of the number of prescriptions per person of all cholesterol-lowering drugs in the NHS in Scotland, in 2024 to 2025.

```{r}
grouped_cl_prescriptions_plot <- grouped_cl_prescriptions %>% 
  mutate(hover_text = paste0("Drug: ", drug_name, "\n",
      "Quantity per Head: ", sprintf("%.2f", total_quantity_per_head))) %>%
ggplot(aes(x = drug_name, y = total_quantity_per_head, fill = drug_name, text = hover_text)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), text = "Times New Roman") +
  theme_custom1() + # customised theme
  scale_fill_manual(values = c(drug_colors)) + #using vector
  facet_wrap(~hb_name) + #facetted plot by HB name
  labs(title = "Number of prescriptions of cholesterol lowering drugs per head in Health Boards in Scotland in 2025", x = "Drug Group", y = "Quantity per Head", fill = "Drug Group")

grouped_cl_prescriptions_plotly <- ggplotly(
  grouped_cl_prescriptions_plot, tooltip = "text") %>% # using plotly so when hovering, we can see the values of the bars
  layout(plot_bgcolor = "#ffffe0") 
grouped_cl_prescriptions_plotly
```

It is clear that four cholesterol-lowering drugs stand out the most in the above plot. We found that atorvastatin is the most prescribed cholesterol-lowering drug in Scotland by the NHS in 2024 and 2025, followed by simvastatin, rosuvastatin, and pravastatin. This finding is supported by 2019 findings, reporting these to be the most used statin drugs, globally (Lippi et al., 2019). Atorvastatin is often the first to be prescribed, and can be frequently prescribed for primary, secondary, and tertiarty prevention of coronary heart disease (McIver, 2020). We will explore atorvastatin, simvastatin, pravastatin, and rosuvastatin only, as these overwhelmingly dominate the cholesterol-lowering drug prescription trends in Scotland, as seen in Plot 2.

# Comparing SE data with cholesterol-lowering prescriptions: Plot 3

We aim to investigate the correlation or relationship between the prescription of the four most prescribed cholesterol-lowering drugs in each health board, with the percentage of households in the upper socioeconomic percentile of that household, using a multi-layered plot.

## Data Wrangling

```{r}
# filtering for the 4 most prescribed statins and using mutate to add this as a new column.
four_statins <- grouped_cl_prescriptions %>% 
  filter(str_detect(drug_name, "ATORVASTATIN|SIMVASTATIN|PRAVASTATIN|ROSUVASTATIN")) %>% 
  mutate('drug_name' = case_when(str_starts(drug_name, "ATORVA") ~ "ATORVASTATIN",
                               str_starts(drug_name, "SIMVA") ~ "SIMVASTATIN",
                               str_starts(drug_name, "PRAVA") ~ "PRAVASTATIN",
                               str_starts(drug_name, "ROSUVA") ~ "ROSUVASTATIN", TRUE ~ drug_name))

```

```{r}
# joining four statins data with socioeconomic data
joined_se_4cl_data <- four_statins %>% 
  full_join(se_data_upper, by = c("hb_name")) 
```

## Plotting number of prescriptions and percentage of households in upper socioeconomic group

Plot 3 investigates the hypothesis, whether there is a relationship between percentage of households in the upper SE group, and number of cholesterol-lowering drug prescriptions.

```{r}
line_plot <- joined_se_4cl_data %>% 
ggplot(aes(x = hb_name)) +
  geom_col(aes(y = total_quantity_per_head, fill = drug_name, text = paste0("Drug: ", drug_name, "\n", "Quantity per Head: ", round(total_quantity_per_head, 2)))) + #rounds to 2 decimal places 
  geom_line(aes(y = upper_household_pct, group = 1, text = paste0("Health Board: ", hb_name, "\n", "% in Upper SE Percentile: ", round(upper_household_pct, 2), "%")), color = "black", size = 0.75) +
  scale_y_continuous(name = "Total Quantity per Head", sec.axis = sec_axis(~., name = "% in Upper SE Percentile")) + # creating two axes on the same plot
  scale_fill_manual(values = c(drug_colors)) + # using vector for drug colour scheme
  labs(title = "The correlation between total statin prescription per head and % of population in upper SE percentile in each health board",
       x = "Health Board", 
       fill = "Drug Name") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + #adjusts axis labels 
  theme_custom1() # customised theme

# interactive plot
line_plot_plotly <- plotly::ggplotly(line_plot, tooltip = "text") %>% # hover feature
   layout(plot_bgcolor = "#ffffe0")
line_plot_plotly
```

The British Heart Foundation explains the correlation between deprivation in Scotland and heart disease. In Scotland, those living in the most deprived areas are almost twice as likely to die from heart and cardiovascular disease before age 75 (Heart health inequalities, 2025). This graph shows the correlation between socioeconomic group and number of cholesterol-lowering prescriptions per health board. It is clear that NHS Dumfries and Galloway has the highest number of total prescriptions per person (56.68), with the lowest percentage of the population in the upper socioeconomic group (9.71%). Additionally, NHS Lothian has one of the lowest number of prescriptions (44.21) but the highest percentage of the population in the upper percentile (20.15%). Notably, NHS Grampian yields a similar result to NHS Lothian, whereby the the total number of prescriptions is the lowest, at 43.61, and the percentage in the upper SE percentile is the second highest at 15%.

# Data collected on SE classification and cholesterol-lowering prescriptions: Table 1

## Writing a function for table theme

```{r}
theme_custom2 <- function(joined_se_4cl_data, ...){
  joined_se_4cl_data %>% 
    tab_options(table.background.color = "#ffffe0",
                column_labels.background.color = "indianred",
                row_group.background.color = "sandybrown",
                table.font.names = "Times New Roman")
}
```

## Gtplot of total number of cholesterol-lowering prescriptions and percentage of households in the upper socioeconomic percentile per health board

This table allows for the direct comparison between percentage of households in upper SE group, with total number of prescriptions of atorvastatin, pravastatin, rosuvastatin, and simvastatin combined.

```{r}
#create table
joined_se_4cl_data %>% 
  mutate(hb_name_plus_pct = paste0(hb_name, " (Households in Upper SE group: ", round(upper_household_pct, 2), "%)")) %>% 
  select(-hb_name, -upper_household_pct) %>% 
  group_by(hb_name_plus_pct) %>% 
  gt() %>% 
  theme_custom2() %>% 
  cols_label(drug_name = "Statin Drug Name",
             total_quantity_per_head = "Prescription Quantity per Head") %>% 
  fmt_number(columns = c(total_quantity_per_head), decimals = 2) %>% 
  cols_align(align = "center",
             columns = c(total_quantity_per_head)) %>% 
  tab_header(title = "Quantity per Head of Statin Prescriptions Across Scotland with Percentage of Population in Upper Socioeconomic Group",
             subtitle = "Among Health Boards in Scotland from July 2024 - June 2025") %>% 
  summary_rows(groups = everything(), columns = total_quantity_per_head, fns = list(total = "sum"), fmt = list(~ fmt_number(., decimals = 2))) %>% 
  tab_source_note(source_note = md("[Click here to access the full dataset](https://www.opendata.nhs.scot/dataset/prescriptions-in-the-community)")) %>% 
  tab_source_note(source_note = md("[Here for Scotland census data on population](https://www.scotlandscensus.gov.uk/webapi/jsf/tableView/tableView.xhtml)")) %>% 
  tab_source_note(source_note = md("[Here for Scotland census data on National Statistics NS-SeC per Household Reference Person](https://www.scotlandscensus.gov.uk/webapi/jsf/tableView/tableView.xhtml)"))
#clickable link takes you to the Open NHS website to access full dataset
```

# Conclusion

From this data, we can conclude that there is a sufficient correlation between number of statin prescriptions and percentage of population in the upper socioeconomic group in health boards across Scotland in 2024-2025. This data supports literature stating that factors such as increased BMI, smoking status, and emotional stress are more prevalent in deprived areas, and are therefore more likely to lead to heart disease or cardiovascular complications, requiring cholesterol-lowering drugs. 

Given that Scotland is predicted to have more old individuals and fewer young individuals by 2045 compared to 2020, it indicates that this data are likely to change, with the overall number of cholesterol-lowering prescriptions increasing (Scottish Government, 2022). Further investigations in this data could compare age demography in each health board in Scotland to investigate the potential correlation, whereby this may have a large affect on prescriptions in the future. With that, the data provided show the current status of Scotland and the trends of cholesterol-lowering prescriptions in correlation with socioeconomic group.

# References

Heart health inequalities (2025). Heart health inequalities. [online] British Heart Foundation. Available at: https://www.bhf.org.uk/what-we-do/our-research/heart-statistics/health-inequalities-research/ [Accessed 22 Nov. 2025]. 

Jones, N.R.V., Fischbacher, C.M., Guthrie, B., Leese, G., Lindsay, R.S., McKnight, J.A., Pearson, D., Philip, S., Sattar, N. and Wild, S.H. (2014). Factors associated with statin treatment for the primary prevention of cardiovascular disease in people within 2 years following diagnosis of diabetes in Scotland, 2006-2008. Diabetic Medicine, [online] 31(6), pp.640–646. doi:https://doi.org/10.1111/dme.12409.

Kimenai, D.M., Pirondini, L., Gregson, J., Prieto, D., Pocock, S.J., Perel, P., Hamilton, T., Welsh, P., Campbell, A., Porteous, D.J., Hayward, C., Sattar, N., Mills, N.L. and Shah, A.S.V. (2022). Socioeconomic Deprivation: An Important, Largely Unrecognized Risk Factor in Primary Prevention of Cardiovascular Disease. Circulation, 146(3), pp.240–248. doi:https://doi.org/10.1161/circulationaha.122.060042.

Lippi, G., Mattiuzzi, C. and Cervellin, G. (2019). Statins popularity: A global picture. British Journal of Clinical Pharmacology, [online] 85(7), pp.1614–1615. doi:https://doi.org/10.1111/bcp.13944.

McIver, L.A. and Siddique, M.S. (2020). Atorvastatin. [online] Nih.gov. Available at: https://www.ncbi.nlm.nih.gov/books/NBK430779/. 

Scottish Government (2022) Scotland’s health and demographic profile: National Care Service – Scotland’s health and demographic profile. Edinburgh: Scottish Government. doi: https://www.gov.scot/binaries/content/documents/govscot/publications/research-and-analysis/2022/06/national-care-service-scotlands-health-demographic-profile/documents/scotlands-health-demographic-profile/scotlands-health-demographic-profile/govscot%3Adocument/scotlands-health-demographic-profile.pdf. 










